---
title: "Random Coefficients Modeling DEMO"
author: "Patrick E. McKnight, Ph.D."
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    number_sections: false
    toc_float: true
    df_print: paged
    code_folding: show
  pdf_document: default
---

```{r preamble, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	background = "#ADD8E6",
	cache = TRUE,
	tidy = 'styler',
	class.source = "bg-warning",
	class.output = "bg-success",
	fig_caption = TRUE,
	digits = 3
)
```

```{r libraries, include=FALSE}
using<-function(...) {
    libs<-unlist(list(...))
    req<-unlist(lapply(libs,require,character.only=TRUE))
    need<-libs[req==FALSE]
    if(length(need)>0){ 
        install.packages(need)
        lapply(need,require,character.only=TRUE)
    }
}
using("tidyr","reshape2","plyr","ggplot2","psych","corrplot","DT","formatR","ggExtra","car","lavaan","dagitty","lavaanPlot","lme4","lmerTest")

round_df <- function(df, digits) {
  nums <- vapply(df, is.numeric, FUN.VALUE = logical(1))
  df[,nums] <- round(df[,nums], digits = digits)
  return(df)
}
```

# The Data and Models

The following data come from the now tainted top 50 home run hitters
of all time in Major League Baseball (MLB).  I say tainted because MLB
underwent a substantial steroids scandal that likely began in the mid
1990s and continued until the recent crackdown last year.  The data
from the top 50 home run hitters is relevant to our discussion because
steroids provides a boost to power enabling the player to hit more
home runs.  That boost ought to provide more home runs than would be
expected by year for each player.  


```{r DataPrep}
HR <- read.table("./HR.dat",header=T)

## Create new steroids variable here
HR$steroids <- -1
HR$steroids[HR$Yr > 1993] <- 1 # only players from the steroid era would likely be on the juice
contrasts(as.factor(HR$steroids)) # default dummy coded

## see?  No matter what you feed it, R will convert to a dummy coded variable
## Let's use more sensible names and lables since we will treat steroids as a factor

HR$steroids <- "Innocent"
HR$steroids[HR$Yr > 1993] <- "Suspect"
HR$steroids.f <- as.factor(HR$steroids)

## we now take a break to offer you students a nice resource:
## https://bookdown.org/pingapang9/linear_models_bookdown/chap-contrasts.html
## read it.  Avail yourselves of a fine education for free!

contrasts(HR$steroids.f) # dummy coded

## change to effects coding
contrasts(HR$steroids.f) <- contr.sum

## change to weighted effects coding
#contrasts(HR$steroids.f) <- contr.wsum

tmp.mat <- contrasts(HR$steroids.f)
tmp.mat
table(HR$steroids)

HR$steroids.we <- HR$steroids.f
contrasts(HR$steroids.we) <- c(1,tmp.mat[2,1]*table(HR$steroids)[[1]]/table(HR$steroids)[[2]])
contrasts(HR$steroids.we)

HR$Year1 <- HR$Yr
HR$Year2 <- HR$Yr

split(HR$Year1,HR$Player) <- lapply(split(HR$Yr,HR$Player), min)
split(HR$Year2,HR$Player) <- lapply(split(HR$Yr,HR$Player), max)
HR$Year <- HR$Yr - HR$Year1 + 1
HR$Year.back <- HR$Year2 - HR$Yr + 1
HR$Year.c <- HR$Year - mean(HR$Year)
HR.4 <- subset(HR,Year < 5)
HR.last5 <- subset(HR,Year.back < 6)
HR.last5$years <- abs(HR.last5$Year.back - 6)
```


```{r models}
# ns model below
lm1 <- lm(HR~Year.c*steroids.we,data=HR.last5)
summary(lm1)

lm1.err <- lm(HR~Year*as.factor(steroids), data=HR.last5)
summary(lm1.err)
```
  
What do these results suggest?  The results indicate that players on
steroids performed differently than players not on steroids.  How
differently?  Look at Figure for the main effects and
interaction.

```{r}
plot(HR~Year.c,data=HR.last5, main="Last 5 Years in MLB")
abline(lm(HR~Year.c,data=HR.last5),lwd=2,lty=2,col="red")
abline(lm(HR~Year.c,data=subset(HR.last5,steroids=="Suspect")),lwd=2,col="blue")
abline(lm(HR~Year.c,data=subset(HR.last5,steroids=="Innocent")),lwd=2,col="green")
legend("topright",c("All Data","Steroid Suspects","non-Steroid Suspects"),lty=c(2,1,1),lwd=2,col=c("red","blue","green"))
```

Note that the steroid users (er, suspects) appear to fall off more rapidly than the
non-steroid users (or suspects).  To better understand the
nature of this effect, we ought to decompose the effects by hand.  I
will work out the simple slope and simple intercept on the board so
you can follow each step.

# Looking deeper

## Plots by Player

```{r PlotByPlayer}
ggplot(HR.last5,aes(x=Year,y=HR,col=Player)) + geom_smooth(method="lm",formula=y~x,se=F) + ylim(0,78) + theme(legend.position="none")
```

## Models Anew!

```{r LMmodeling}
lm0 <- lm(HR~1, data=HR.last5)
summary(lm0)
lm1 <- lm(HR~Year.c*steroids.we, data=HR.last5)
summary(lm1)
lme0 <- lmer(HR~ 1 + (1 | Player), data=HR.last5)
summary(lme0)
lme1 <- lmer(HR~Year.c + (1 | Player), data=HR.last5)
summary(lme1)
## lme1 last model that runs (4/20/2023)
lme2 <- lmer(HR~Year.c + (Year.c | Player), data=HR.last5)
## whoopsie!  We met the limits to our data and modeling.  Let's try a different attack
lme3 <- lmer(HR~Year.c + steroids.we + (1 | Player), data=HR.last5)
summary(lme3)
lme4 <- lmer(HR~Year.c + steroids.we + (steroids.we | Player), data=HR.last5)
## whoopsie!  We met another limit.  New attack
lme5 <- lmer(HR~Year.c*steroids.we + (1 | Player), data=HR.last5)
summary(lme5)
```

## Coeficients

```{r LMERcoefs}
round(coef(lm0),2)
round(coef(lme0)$Player,2)
round(coef(lme1)$Player,2)
round(coef(lme3)$Player,2)
round(coef(lme5)$Player,2)
```


```{r nestMcomp}
anova(lme0,lme1,lme3,lme5)
```

